---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
# IBM HR Analytics - Employee Attrition 

*by Ankit Naik*
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(corrplot)
library(ggthemes)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv("~/Downloads/IBM-HR.csv")
```

## Abstract

In this document I will describe how various factors affect attrition rate in
IBM and find out factors which can help reducing attrition rate by avoiding some
aspects which is explainatory.

## Introduction

IBM provides pseudo-dataset that shows key factors from which we can see 
patterns regarding performance and attrition of an employee. This notebook will 
focus on finding out factors which directly affects attrition rate.

I will proceed to analysis with some common questions in target like -

**Whether no promotion leads to attrition ?**

**Whether less income leads to attrition ?**

**Is the person leaves job frequently ?**

and many more.

## Univariate Plots

Lets start by exploring the dataset by summarizing it - 

```{r echo=FALSE, Univariate_Plots}

str(df)

```

Dropping Standard Hours and Employee count from dataframe as its un-necessary
fot our analysis.

```{r}
df$StandardHours <- NULL
df$EmployeeCount <- NULL
```


Checking if there are any missing values in dataset by checking dimension - 

```{r echo=FALSE}
dim(df)
```

Since dimension is same as the values in summary **there are no missing values**. 
Hence we can continue with our exploration.

Lets explore demographics of our datset like Age,Gender,Attrition and Distance 
From Home

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes(x = Age),data = df) + geom_bar(binwidth = 5) + 
  ggtitle("Age Distribution")

p2 <- ggplot(aes(x = Gender),data = df) + geom_bar() + 
  ggtitle("Gender Distribution")

p3 <- ggplot(aes(x = Attrition) , data = df) + geom_bar()+ 
  ggtitle("Attrition Distribution")

p4 <- ggplot(aes(x = DistanceFromHome) , data = df) + geom_bar()+
  ggtitle("DistanceFromHome Distribution") + scale_y_log10()


grid.arrange(p1,p2,p3,p4, ncol = 2)

```

*Distribution of Age* - Age is normally distributed with mean age around 35 and 
minimum age being 18 and maximum bieng 60 (Retirement). This suggests there are
no unusual data in dataset.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$Age)
```


*Distribution of Gender* - Female are less than Male which is not unusual in 
workplace. While comparing something wrt Gender we need to take care of this.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$Gender)
```

*Distribution of Attrition* - This is also normal since attrition is always very
less in a well established company.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$Attrition)
```

*Distribution of Distance From Home* - This data is interesting to look as this 
follows common idea of living near the office. Interesting part will be
If someone leaves company due to this factor.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$DistanceFromHome)
```


Additionally lets explore some more features - 

```{r echo=FALSE}
p33 <- ggplot(aes(x = BusinessTravel),data = df) + geom_bar() + 
  ggtitle("Travel Distribution")

p34 <- ggplot(aes(x = JobLevel),data = df) + geom_bar() + 
  ggtitle("Job Level Distribution")

p35 <- ggplot(aes(x = TotalWorkingYears) , data = df) + geom_bar()+ 
  ggtitle("Experience Distribution") + scale_y_log10()

p36 <- ggplot(aes(x = MonthlyIncome) , data = df) + geom_histogram()+
  ggtitle("Income Distribution")


grid.arrange(p33,p34,p35,p36, ncol = 2)
```

*Distribution of Travel* - Graph suggests most of the employee are from travel
rarely job type. This will be a good feature to see attrition rate.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$BusinessTravel)
```

*Distrubution of Job Level* - Dataset contains more people from lower level job
which is normal hence not many people are present in higher level job.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$JobLevel)
```

*Distribution of Experience* - Similar to job level low experience employee are 
more because of the same reason explained above.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$TotalWorkingYears)
```
*Distribution of Income* - This is also related to age,experience and job level 
hence same pattern can be observed here.

Lets summarise this feature to explain findings - 

```{r echo=FALSE}
summary(df$MonthlyIncome)
```

All this relation can be explored from correlation plot which I will be doing
in bivariate plot section.



## Univariate Analysis

##### What is the structure of dataset?

The dataset includes all details of employee with some factor variables like
Attrition,Business Travel,Department,etc. Others are Integer like distance from 
home, Monthly Income, etc. There are no other datatypes and no missing values.

##### What is/are the main feature(s) of interest in dataset?

Some interesting variables that definitely will help in finding out factors 
affecting attrition are one and foremost Attrition, others are Age, Business 
Travel, Distance From Home, Monthly Income, Overtime and Work-Life Balance

##### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Features like time since last promotion, Companies worked, Gender and Maritial 
Status will help exploring in deep.

# Bivariate Plots Section

Before starting bivariate plots, lets compare correlation between various 
components such that we can reduce unnecessary comparision between related 
components and focus on important variables.

```{r echo=FALSE, Bivariate_Plots}

nums <- sapply(df, is.numeric)

int_data <- df[,nums]
int_attrition <- as.numeric(df$Attrition) - 1 
int_data <- cbind(int_data,int_attrition)

m1 <- cor(int_data)

corrplot(m1,method = 'circle',tl.cex = 0.5,type = "upper",diag=FALSE,
                                              title = "Correlation Plot") 
```

From above correlation graph we can deduce that following data are strongly 
correlated (Dark blue color represents high correlatoion)-

**1. Age ~ JobLevel ~ MonthlyIncome ~ TotalWorkingYears ~ YearsAtCompany** - 

This makes sense as age increases Salary, Job Level and Monthly Income
increases. Lets use scatter plot to plot all these variables to look in detail.

```{r echo=FALSE, message=FALSE, warning=FALSE}

p5 <- ggplot(aes(x = JobLevel , y = Age) , data = df) +
  geom_jitter(alpha = 0.3) +
  ggtitle("Joblevel V/S Age")

p6 <- ggplot(aes(x = MonthlyIncome , y = Age) , data = df) +
  geom_jitter(alpha = 0.5) +
  ggtitle("MonthlyIncome V/S Age") + 
  xlab("Monthly Income (in Dollars)")

p7 <- ggplot(aes(x = TotalWorkingYears , y = Age) , data = df) + 
  geom_jitter(alpha = 0.5) +
  ggtitle("TotalWorkingYears V/S Age")

p <- ggplot(aes(x = YearsAtCompany , y = Age) , data = df) + 
  geom_jitter(alpha = 0.5) + 
  ggtitle("YearsAtCompany V/S Age")

grid.arrange(p5,p6,p7,p,ncol=2)

```

As from the graph we can see all these features increases as age increase.

**2. JobLevel ~ YearsAtCompany ~ TotalWorkingYears**

This is interesting relation, as job level increases people remain in the same
company. Lets see with scatter plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
p8 <- ggplot(aes(x = JobLevel , y = YearsAtCompany) , data=df) + 
  geom_jitter(alpha = 0.3)+ 
  ggtitle("JobLevel V/S YearsAtCompany")
p9 <- ggplot(aes(x = JobLevel , y = TotalWorkingYears) , data=df) + 
  geom_jitter(alpha = 0.3) + 
  ggtitle("JobLevel V/S TotalWorkingYears")

grid.arrange(p8,p9 , ncol = 2)

```

One strange finding here is job level is not so dependent factor with Years
at company as compared to total working years. People working in same companies
has different level with same age while this is not true for total experience.

**3. PercentSalaryHike ~ PerformanceRating**

As performance rating improves hike is imminent, this relationship can be seen 
here as increase in rating gives more hike.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = PerformanceRating , y = PercentSalaryHike) , data=df) + 
  geom_jitter(alpha = 0.5) + ggtitle("Rating V/S Hike") 

```

Since dataset contains only rating of 3 and 4. We can easily conclude people 
with rating 3-3.5 get less than 20% hike and people with rating 3.5-4 get more 
than 20% hike.

**4. YearsAtCompany ~ YearsInCurrentRole ~ YearsSinceLastPromotion ~ YearsWithCurrManager**

This has shown some interesting relation like years at company and promotion.
With increase in experience at same company primoition does not happen as usual 
and role is also not getting changed frequently. These two factor contribute to 
longer time with current manager.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p11 <- ggplot(aes(x = YearsInCurrentRole , y = YearsAtCompany) , data=df) +
  geom_jitter(alpha = 0.2)
p12 <- ggplot(aes(x = YearsSinceLastPromotion , y = YearsAtCompany) , data=df) + 
  geom_jitter(alpha = 0.2)
p13 <- ggplot(aes(x = YearsWithCurrManager , y = YearsAtCompany) , data=df) + 
  geom_jitter(alpha = 0.2)

grid.arrange(p11,p12,p13,ncol =3)
```


Now after exploring relationships with dataset we can start with our posed 
question. **Which factors attrition depends on?**

As there is no direct relation of attrition with other values lets start looking
into factors one by one-


##### Attrition based on Age

Age is one of the major factor attrition depends on. Young people tends to switch
more job than older people. Lets explore the pattern. From here on I will be 
producing plots in proportion. Proportion will be comparable to actual variable 
and attrition at that variable.

Plotting histogram of Age will answer most of the questions since age is related
to numerous factors in dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=Age,..count.., fill = Attrition) ,  data = df) + 
  geom_bar(position = "fill") +
  scale_fill_excel() + theme_excel() + 
  scale_x_continuous(breaks = seq(18,62,2)) + ylab("Mean") +
  ggtitle("Distribution of attrition year by year")

```
From the graph above it is clear that people who joined company at young age 
tend to leave the company either for higher education or for other company.
The pattern reduces over time and after 50 years it starts increasing again 
due to factors like retirement.

Since age and total working years are related we can see the same pattern here. 
With start of career attrition is more and it reduces afterwards. At age near to 
retirement around 40 Attrition rate increases exponentially. Same goes for 
JobLevel and Monthly Income.

Lets plot some more variable based on attrition

##### Attrition based on Business Travel and Year in Current Role

```{r echo=FALSE, message=FALSE, warning=FALSE}
p14 <- ggplot(aes(x = BusinessTravel,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() + 
  ylab("Mean") +  
  theme(axis.text.x = element_text(size  = 10, angle = 45,hjust = 1,vjust = 1))

p17 <- ggplot(aes(x = YearsInCurrentRole,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() + 
  ylab("Mean")


grid.arrange(p14,p17,ncol = 2)
```

As expected people who have to travel frequently has higher attrition.

Same trend is followed in Years at company initial attrition is higher and after
23 years attrition rate begins to increase. Which can amount to people taking 
retirement.

Years in current role which may change after promotion usually shows either 
people leave just after getting promoted or they leave because they are not 
getting new role. But pattern is non explainatory

##### Attrition based on Year Since last Promotion,Work-Life Balance,Stock\ 
Option Level and Percent Salary Hike

```{r echo=FALSE, message=FALSE, warning=FALSE}
p18 <- ggplot(aes(x = YearsSinceLastPromotion,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p19 <- ggplot(aes(x = WorkLifeBalance,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p20 <- ggplot(aes(x = StockOptionLevel,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p21 <- ggplot(aes(x = PercentSalaryHike,..count.. ),data = df) +
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() + 
  ylab("Mean")


grid.arrange(p18,p19,p20,p21,ncol = 2)
```

Popular opinion will suggest that years since last promotion matters for 
attrition but based on our data its hard to deduce that not getting promoted is 
reason for attrition. Also IBM employee hierarchy is different then other 
companies which has fewer job profile hence less promotion but pay increases
overtime.

Work-Life balance is a factor in attrition as people reporting low work-life 
balance have very high attrition compared to others. Same goes for stock option 
level, lower level leave and higher level attrition can be due to people retiring 

Again percent salary hike shows that people getting more hike leave after the 
hike. Which is true as other companies will try to match up the salary given in 
the previos job. But difference is not large enough to conclude this.

##### Attrition based on Overtime,No. of companies Worked,Maritial Status and Job Satisfaction

```{r echo=FALSE, message=FALSE, warning=FALSE}
p22 <- ggplot(aes(x = OverTime,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") + 
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p23 <- ggplot(aes(x = NumCompaniesWorked,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill",binwidth = 2)+
  scale_fill_excel() + theme_excel() + 
  ylab("Mean")

p24 <- ggplot(aes(x = MaritalStatus,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() + 
  ylab("Mean")

p25 <- ggplot(aes(x = JobSatisfaction,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() + 
  ylab("Mean")

grid.arrange(p22,p23,p24,p25,ncol = 2)
```

As its clear that overtime does have an affect on attrition as people doing 
overtime tends to leave company.

Interestingly people worked with more than 5 companies does have more attrition 
than others. But pattern is not clear hence we cannot conclude.

As expected people with MaritialStatus single has more attrition as after getting
married leaving job is very rare since supporting family is an issue.

Job satisfaction is also an issue as people with low job satisfaction leave the 
job.

##### Attrition based on Job Involvement,Environment Satisfaction,EducationField\
and Education

```{r echo=FALSE, message=FALSE, warning=FALSE}
p26 <- ggplot(aes(x = JobInvolvement,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") +
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p27 <- ggplot(aes(x = EnvironmentSatisfaction,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")

p28 <- ggplot(aes(x = EducationField,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")+
  theme(axis.text.x = element_text(size  = 10, angle = 45,hjust = 1,vjust = 1))

p29 <- ggplot(aes(x = Education,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() +
  ylab("Mean")

grid.arrange(p26,p27,p28,p29,ncol = 2)
```

Again job envolvement and environment satisfaction is directly affecting 
attrition.

HR, Marketing and Technical Degree does have more attrition than any other field.

Attrition does not depend much on Education but we can say that people with 
highest level of education have low attrition.

##### Attrition based on Distance From Home and Department

```{r echo=FALSE, message=FALSE, warning=FALSE}
p30 <- ggplot(aes(x = DistanceFromHome,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill") +
  scale_fill_excel() + theme_excel() +  
  ylab("Mean") 

p31 <- ggplot(aes(x = Department,..count.. ),data = df) + 
  geom_bar(aes(fill=Attrition),position = "fill")+
  scale_fill_excel() + theme_excel() +  
  ylab("Mean")+
  theme(axis.text.x = element_text(size  = 10, angle = 45,hjust = 1,vjust = 1))

grid.arrange(p30,p31,ncol = 2)
```

Distance from home may not look like a factor for attrition but people living 
in larger distance from office does have more attrition.

It may look like department effects attrition but there are no clear 
patterns to conclude that.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Most interesting feature was age, its directly related to other features and 
exploring only age answered most of the questions like less income, job level
and new joinee has more attrition than others.

Other feature which was good to look at was Overtime. Attrition rate for employee
having overtime was much more compared to other employee.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Relationship between Years at company and Last promotion was interesting as 
More years at company promotion was not happening as usual. Which is contarary
to popular belief. Also role was not getting changed.


### What was the strongest relationship you found?

Strongest relationship was between Age an Monthly Income. As age increased, 
income of an individual increases. 


# Multivariate Plots Section


##### Attrition based on Travel and Age

Idea of this multivariate plot was to see if increase in age and travel are
related factor for attrition. Lets see with plot


```{r echo=FALSE, Multivariate_Plots}

ggplot(aes(x = BusinessTravel  , y = Age) , data = df) + 
  geom_jitter(aes(color=Attrition),alpha = 0.8) +  
  scale_fill_excel() + theme_dark() + 
  ggtitle("Comparision of Business Travel and Age based on attrition")

```

Although attrition in non-travel job type is very low but above age 40 attrition
is almost nill. While other travel type still sees some attrition. Another 
conclusion we can give is for travel-frequently below age 25 attrition is almost
100%.

##### Attrition based on Business Travel and Maritial Status

Two feature that I wanted to see that if someone is married and there is travel 
required. Will people leave the job. Lets look at plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = BusinessTravel  , y = MaritalStatus) , data = df) + 
  geom_jitter(aes(color=Attrition),alpha = 0.8) + 
  scale_fill_excel() + theme_dark() +  
  ggtitle("Comparision of Business Travel and Maritial Status based on attrition")
```

After looking at plot we can definitely say there are no clear pattern visible 
here. Hence there is no effect of this reason on attrition.

##### Attrition based on Distance and Age with Maritial Status

Since distance and age can be a factor on leaving job as people get tired 
travelling also adding to factor is if someone has a family. Lets look at plot.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = DistanceFromHome  , y = Age) , data = df) + 
  facet_grid(~ MaritalStatus) +
  geom_jitter(aes(color=Attrition),alpha = 0.8) + 
  scale_fill_excel() + theme_dark() + 
  ggtitle("Comparision of Distance from home, age and Maritial Status based on 
          attrition")
```

Here also there are no clear pattern hence assumption is false.

##### Attrition based on Travel and Gender

Gender based analysis can give some clear result about weather gender is a 
factor on attrition due to travel.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = BusinessTravel  , y = Gender) , data = df) + 
  geom_jitter(aes(color=Attrition),alpha = 0.8) + 
  scale_fill_excel() + theme_dark() + 
  ggtitle("Comparision of Gender and Business Travel based on Attrition ")
```

Here also we can see gender is not that big factor as there is no clear pattern.

##### Attrition based on Distance,Gender and Maritial Status


Adding one more variable Marritial Status for above plot. Lets explore,

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = Gender  , y = DistanceFromHome) , data = df) + 
  facet_grid(~MaritalStatus) + 
  geom_jitter(aes(color=df$Attrition),alpha = 0.8) + 
  scale_fill_excel() + theme_dark() + 
  ggtitle("Comparision of Distance from home and Gender based on attrition")
```

Here we can see some pattern that Married people has more attrition at larger
distance compare to single and divorced.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

One definitive relation was Business Travel does play role in attrition as 
Travel Frequently has more attrition at age less than 25 and Non travel job
has almost no attrition after age 40. While there is no effect of travel rarely
job on attrition.

Other relationship was married people have more attrition when distance from home
is large compare to other.

### Were there any interesting or surprising interactions between features?

Surprising interaction is Gender plays no role in attrition when compared to
any of the feature in dataset. While popular opinion suggests otherwise.

------

# Final Plots and Summary


### Plot One



```{r echo=FALSE, Plot_One}
ggplot(aes(x = TotalWorkingYears) , data = df) + geom_bar()+ 
  ggtitle("Experience Distribution") + scale_y_log10()
```

### Description One

Reason behind including this plot in summary was the increasing pattern of 
experience till 10 years and reducing after that which suggests most of the
employee start retiring after working 10 years and 40 is the max limit which is
Retirement age.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = Age,..count.. ),data = df) + 
    geom_bar(aes(fill=Attrition),position = "fill") +  
    scale_fill_excel() + theme_excel() + 
    ggtitle("Distribution of attrition year by year based on age")
```


### Description Two

Age is one of the feature which related to several other features hence we were
able to directly deduce feature like Monthly Income, years at company , etc has
its effect in attrition in same pattern related to age.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = BusinessTravel  , y = Age) , data = df) + 
    geom_jitter(aes(color=Attrition),alpha = 0.8) +  scale_fill_excel() + 
    theme_dark() + 
    ggtitle("Comparision of Business Travel and Age for Attrition")
```

### Description Three

Although data points were not equal but from this plot we were able to see that
in Non-Travel job people with age more than 40 had more attrition. Simmilarly 
people travelling frequently had allot of attrition below age 30(almost 80%).

------

# Reflection

Whole analysis was based on finding out which factors affect Attrition. Based
on analysis following factors came up -

######1. Age - Initital Years,
######2. Job Level - Low Job Level
######3. Monthly Income - Low Income
######4. Working Years - Less Working Years
######5. Business Travel - More travel
######6. Stock Option - Low level
######7. Overtime - Yes
######8. Maritial Status - Single
######9. Job Satisfaction - Low
######10. Job Involvement - Low
######11. Environment Satisfaction - Low
######12. Distance From Home - Large
######13. Department - Sales and HR
######14. Travel Frequency - Frequent and Married

One of the problem faced with dataset is its not given what was the reason 
attrition has happened. Including it will make the analysis more exciting.
Also dataset is not enough to conclude some of the things.

Surprisingly attrition is happening for the same reason as percieved. But the
finding that attrition is not related to promotion was great finding in the 
analysis..

Future work on this can be to provide a model where we can give details how to 
reduce attrition by avoiding some factors.

